---
title: "SDS291 Group Project - Final Proposal"
author: "Tseegi Nyamdorj, Tina Chang, Jinwen Xiang, My My Tran"
date: '2023-05-08'
output: 
  pdf_document:
    toc: yes
    toc_depth: '2'
  html_document:
    toc: yes
    toc_depth: 2
    toc_float: yes
    fig_width: 7
    fig_height: 6
    fig_caption: yes
    theme: lumen
    df_print: paged
    code_folding: show
---
# Title: The Influence of Parenthood and Age on Adults Pursuing Higher Education

# Abstract

The purpose of our analysis is to explore factors that influence the likelihood of an adult pursuing higher education. Using data from the National Surveys of College Graduates (NSCG), Recent College Graduates (NSRCG), and Doctorate Recipients (SDR) available in the IPUMS database, we explore the likelihood of an adult to pursue higher education is based on the number of children the adult has, as well as their age. We will compare three different logistic regression models with enrolled/graduated higher education in the last five years as the response variable. Through our research, we have concluded that a person's age and the number of children are significant enough to contribute to predicting the chances of an adult pursuing higher education. 

# Introduction

The aim of this project is to investigate the likelihood of adults obtaining higher education, with a specific focus on adults with children and the number of children they have, and the age of the adult in the United States. Previous research suggests that having children may affect a parent's pursuit of higher education. For instance, Kulp (2016) found that "mothers with doctorates tended to take longer to complete the degree and graduated at an older age than their non-parent counterparts ^[Catalano, Amy J; Radin, Susan T. International Journal of Doctoral Studies; Santa Rosa. Parents Pursuing a Doctorate of Education: A Mixed Methods Examination of How Parents Manage the Roles of Student and Parent. Vol. 16,  (2021): 253-272. DOI:10.28945/4741]. We find this topic to be particularly interesting because of the Ada Comstock program at Smith College, allowing non-traditional aged students to pursue higher education and understanding the significance that age and number of children have on parent adults in higher education. 

Our primary hypothesis is that adults with no children or fewer children are more likely to pursue higher education in their adult years than adults with 3 or more children. Our secondary hypothesis is that older adults are less likely to have attended higher education within the last 5 years than younger adults. We will be comparing logistic models to see which variables, between age and number of children, better predict the likelihood of adults pursuing higher education, meaning they have completed their degree or are part-time or full-time students within the years 2006-2010. 

Additionally, previous studies have shown that student parents tend to be older^[Institute for Women’s Policy Research (IWPR). 2018. Institute for Women’s Policy Research (IWPR) analysis
of data from the U.S. Department of Education, National Center for Education Statistics, 2015–16 National
Postsecondary Student Aid Study (NPSAS:16)] and are more likely to face intense economic challenges^[ Noll, Liz & Reichlin, Lindsey & Gault, Barbara. (2017). College Students with Children: National and Regional Profiles. 10.13140/RG.2.2.21338.75205.] while enrolled in college than their peers without children, however there is no substantive research on the number of children dependent on a student parent. Regardless, through our analysis, we hope to offer more clarity in the effects that each variable has on adults pursuing higher education. 

# Methods

## Data

This study will use data from the National Surveys of College Graduates (NSCG), Recent College Graduates (NSRCG), and Doctorate Recipients (SDR) available in the IPUMS database. The NSCG is a survey of individuals residing in the United States who hold a bachelor's degree or higher, while the SDR collects data from individuals who received a doctorate degree from a US institution. The NSRCG collects data from individuals who obtained a science or engineering degree from a US institution within the last two years. All three surveys collect data from individuals under the age of 76. To increase our sample size, we will be using data from the year 2010, and our total sample size is 108,337 adults. The data is available at this URL: https://highered.ipums.org/highered-action/variables/group?id=demog. 

## Variables
The variables in our analysis include binary response variable of an adult graduating/studying in a higher education in the last five years with primary categorical explanatory variable of number of children, and secondary numerical explanatory variable of adult’s age. 

Outcome variables: Binary variable indicating whether or not the adult is pursuing or has pursued (graduated) from higher education as a part-time/full-time student in the last 6 years (binary: 0 indicates false and 1 indicates true). For analysis, we separated it into two binary values by whether their year of most recent degree falls under the last 6 years and/or is currently a full-time or part-time student. 

Explanatory variables: number of children (categorical), age of parent (numerical) 
Primary explanatory variable is the number of children, which is categorical with three levels of factored labels of: “No Child,” “One Child,” and “2 or More Children.” 
Secondary explanatory variable is the age of the parent, which is numerical from 23 to 75.

## Analysis

We performed two model comparisons between 3 logistic regression models to analyze our primary and secondary hypothesis. The response variable is the binary variable for whether the adult has pursued higher education in the last 6 years.

The \textbf{primary hypothesis} is that adults with no children or fewer children are more likely to pursue higher education in their adult years than adults with 3 or more children.

\textbf{Reduced:}
$logit(degree within last 6 years)=\beta_0+\beta_1 * 1(nochild) +\beta_2 * 1(one child)+\beta_3 * 1(2+ children)$

\textbf{Full:}
$logit(degree within last 6 years)=\beta_0+\beta_1 * 1(nochild) +\beta_2 * 1(one child)+\beta_3 * 1(2+ children) +\beta_4 * AGE$

\textbf{The test of our hypothesis will be:}
$H_0: \beta_3 and \beta_4 = 0$
$H_A: \beta_3 or \beta_4 \neq0$

\textbf{Secondary hypothesis:} Younger adults are more likely to have attended higher education within the last six years than older adults.

\textbf{Reduced:}
$logit(degree within last 6 years) = \beta_0+\beta_1 * AGE$

\textbf{Full:}
$logit(degree within last 6 years) = \beta_0+\beta_1 * 1(nochild) +\beta_2 * 1(one child)+\beta_3 * 1(2+ children) + \beta_4 * AGE$

\textbf{The test of our hypothesis will be:}
$H_0: \beta_1 = \beta_2 = \beta_3 = \beta_4 = 0$
$H_A: \beta_1 or \beta_2 or \beta_3 or \beta_4 \neq0$

\textbf{Results:}
When comparing logistic regression models, we use the chi-squared test. We use this to see if between the two nested models are inconsistent with differences that are expected under the null hypothesis (from class slides on Comparing Logistic Regression Models).

The following are our two model comparisons for our hypotheses:

\textbf{Primary hypothesis: Does the number of children that an adult has contribute to predicting the chances of an adult pursuing or graduating higher education in the last five years?}

In the reduced model, numberChild_6years_simple, the intercept (having no children) is -0.717, the number of children (1) is -0.617 and number of children (3+) is -1.055.

In the full model, model_6years, the intercept (having no children) is 5.922, age is -0.166, the number of children (1) is -0.50, the number of children (3+) is -0.937.

```{r}
library(ipumsr)
library(Stat2Data)
library(ggplot2)
library(dplyr)
library(moderndive) 
```

```{r}
# NOTE: To load data, you must download both the extract's data and the DDI
# and also set the working directory to the folder with these files (or change the path below).

if (!require("ipumsr")) stop("Reading IPUMS data into R requires the ipumsr package. It can be installed using the following command: install.packages('ipumsr')")

ddi <- read_ipums_ddi("highered_00002.xml")
data <- read_ipums_micro(ddi)
```

```{r}
data <- data[,!names(data) %in% c("WEIGHT", "SAMPLE", "SURID")]
```

```{r}
data_6years <- data %>%
  mutate(last06 = case_when(MR03Y5 >= 2006 | ACFPT == 1 | ACFPT == 2 ~ 1, 
                            MR03Y5 < 2006 | ACFPT == 98 | ACFPT == 3 ~ 0)) %>%
  mutate(chU6 = case_when(CH6IN == 0 | CH6IN == 98 ~ 0,
                          CH6IN == 1 ~ 1)) %>%
  mutate(GENDER = GENDER -1)
```
```{r}
data_6years$CHTOT <- factor(data_6years$CHTOT,
                       levels = c(98, 1,3),
                       labels = c("No Child", "One Child", "2 or More Children")
                       )
```

```{r}
numberChild_6years_simple <- glm( last06 ~ CHTOT, data = data_6years, family = binomial)
summary(numberChild_6years_simple)

```
```{r}
model_6years <- glm( last06 ~ AGE + CHTOT, data = data_6years, family = binomial)
summary(model_6years)
```

```{r}
anova(numberChild_6years_simple, model_6years, test="Chisq")
```


Using ANOVA to perform a G-test, (p < 0.05) meaning we reject the null hypothesis and prefer the full model. We conclude that the number of children does contribute to predicting the chances of an adult pursuing or graduating higher education in the last five years.

\textbf{Secondary hypothesis: Does the age of an adult contribute to their chances of pursuing or graduating higher education in the last five years?}

In the reduced model, age_6years_simple, the intercept is 6.031 and the age is -0.179.

In the full model, model_6years, the intercept (having no children) is 5.922, age is -0.166, the number of children (1) is -0.50, the number of children (3+) is -0.937.


```{r}
age_6years_simple <- glm( last06 ~ AGE, data = data_6years, family = binomial)
summary(age_6years_simple)
```
```{r}
model_6years <- glm( last06 ~ AGE + CHTOT, data = data_6years, family = binomial)
summary(model_6years)
```

```{r}
anova(age_6years_simple, model_6years, test="Chisq")

```


Using ANOVA to perform a G-test, (p < 0.05) meaning we reject the null hypothesis and prefer the full model. We conclude that age does contribute to predicting the chances of an adult pursuing or graduating higher education in the last five years for adults, and adults who have children.

\textbf{Coefficient Interpretation:}

For reduced model \emph{last06~AGE:}

The intercept represents that when age is 0, the log-odds of the adult pursuing or has pursued (graduated) from higher education as a part-time/full-time student in the last 6 years is 6.03. 

The AGE coefficient represents that the log-odds of the adult pursuing or has pursued (graduated) from higher education as a part-time/full-time student in the last 6 years decreases by 0.18 when the age variable increases by 1. 


For reduced model \emph{last06 ~ CHTOT:}

The intercept represents that when CHTOT is 0, the log-odds of an adult pursuing or has pursued (graduated) from higher education as a part-time/full-time student in the last 6 years is -0.72.

The CHTOTOne Child represents that compared to the no child level, the log-odds of the adult pursuing or has pursued (graduated) from higher education as a part-time/full-time student in the last 6 years with one child decreases by 0.62 compared to no children.

The CHTOT2 or More Children represents that compared to no child level, the log of the odds of the adult pursuing or has pursued (graduated) from higher education as a part-time/full-time student in the last 6 years with two or more children decreases by 1.05 compared to no children.

For full model \emph{last06 ~ AGE + CHTOT:}

The intercept represents that when a person is 0 years old with no children, the log-odds of the person pursuing or has pursued (graduated) from higher education as a part-time/full-time student in the last 6 years is 5.92. 

The AGE coefficient represents that for each additional age, the log-odds of the adult pursuing or has pursued (graduated) from higher education as a part-time/full-time student in the last 6 years decreases by 0.17.

The CHTOTOne Child represents that compared to the no child level and with age 0, the log-odds of the person pursuing or has pursued (graduated) from higher education as a part-time/full-time student in the last 6 years with one child is 0.5 lower compared to no children.

The CHTOTOne Child represents that compared to the no child level and with age 0, the log-odds of the person pursuing or has pursued (graduated) from higher education as a part-time/full-time student in the last 6 years with one child is 0.94 lower compared to no children.

# Exploratory Visualization


```{r}
probabilities <- model_6years %>% 
  predict(data_6years, type = "response")
predicted.classes <- ifelse(probabilities> 0.5, 0, 1)
mean(predicted.classes == data_6years$last06)
```
```{r}
new_data1 <- exp(cbind(OR = coef(model_6years), confint(model_6years)))
head(new_data1)

```

```{r}
data_6years %>%
  ggplot(aes(x=AGE, y=last06, color= CHTOT)) +
  geom_point(
    position=position_jitter(height=.01)
    ) +
  geom_smooth(
    method=glm,
    method.args=list(family=binomial),
    se=FALSE,
    formula = y~x
    )
```



# Discussion

In the study, we want to investigate how the number of children or the age of the adult would influence the likelihood of adults pursuing higher education in the United States. After analysis, we conclude that both factors (age, number of children) contribute significantly to predicting the chances of adults obtaining higher education in the last five years. To streamline the analysis process, we limited our data to the last five years, and the data we used only includes adults who are younger than 76, meaning we cannot generalize our results to the whole US adult population. Although we conclude that both of these factors have significant contributions to our model, we cannot measure the exact strength of the two variables when doing model comparisons between logistic regression models. 

To improve our study, we could include more data. For instance, instead of only investigating data in the last five years, we can expand our data set to last ten-year data. Also, since the number of children CHTOT variable is categorical with three levels (no children, 1 child, or 3+ children), this variable limited the model we could use. In our study, we used a logistic regression model, but if the CHTOT variable was numerical with a range of ages, we could possibly fit linear regression models to see the relationship between age and the number of children, which could provide more insight into their significance. Additionally, since our data was mainly categorical, it would have been interesting to see if there were significant contributions between other variables such as minority status, and income/salary.

In summary, though there are some limitations, the study provides important insight into the relationship between the number of children, the age of the adult, and the likelihood of adults pursuing higher education.

# References

Catalano, Amy J; Radin, Susan T. International Journal of Doctoral Studies; Santa Rosa. Parents Pursuing a Doctorate of Education: A Mixed Methods Examination of How Parents Manage the Roles of Student and Parent. Vol. 16,  (2021): 253-272. DOI:10.28945/4741.

Institute for Women’s Policy Research (IWPR). 2018. Institute for Women’s Policy Research (IWPR) analysis
of data from the U.S. Department of Education, National Center for Education Statistics, 2015–16 National
Postsecondary Student Aid Study (NPSAS:16)

Noll, Liz & Reichlin, Lindsey & Gault, Barbara. (2017). College Students with Children: National and Regional Profiles. 10.13140/RG.2.2.21338.75205. 
